A Decision Procedure for (Co)datatypes in SMT Solvers

### 2016-02-09 sister/sister.tex
avoid "tableau" terminology at IJCAI
more
more typesetting

@@

Given a fixed signature, %~$\Sig$,
the decision procedure for the universal theory of (co)data\-types %$\thD$
determines the
%$\thD$-
satisfiability of finite sets $\Ec$ of literals:\ equalities
and disequalities between first-order terms, %$\Sig$-terms,
whose variables are interpreted existentially.
The decision procedure is formulated as a tableau-like calculus.
%Following the tradition of refutational theorem proving,
Proving a universal quantifier-free conjecture is reduced to showing that
its negation is unsatisfiable. 
%The presentation is
%inspired by Barrett et al.\ \shortcite{barrett-et-al-2007} but at
%a higher level, using unoriented equations instead of oriented ones.

@@

@@

@@

\begin{figure}[t!]
\small
\centering
\begin{tabular}{c}
\(
\inferrule{
  t \in \tEc
}{
  \Ec := \Ec,\: t \teq t
}
\)
\rn{Refl}
\quad
\(
\inferrule{
 t \teq u \in \Ec
}{
 \Ec := \Ec,\: u \teq t
}
\)
\rn{Sym}
\quad
\(
\inferrule{
  s \teq t,\; t \teq u \in \Ec
}{
  \Ec := \Ec,\: s \teq u
}
\)
\rn{Trans}
\\[5\jot]
\(
\inferrule{
  \vec t \teq \vec u \in \Ec
  \betweenantes
  \const f( \vec t \,),\, \const f( \vec u ) \in \tEc
}{
  \Ec := \Ec,\: \const f( \vec t \,) \teq \const f( \vec u )
}
\)
\rn{Cong}
\quad
\(
\inferrule{
  t \teq u,\; t \tneq u \in \Ec
}{
  \bot
}
\)
\rn{Conflict}
\\[5\jot]
\(
\inferrule{
  \const{C}( \vec t \,) \teq \const{C}( \vec u ) \in \Ec
}{
  \Ec := \Ec,\: \vec t \teq \vec u
}
\)
\rn{Inject}
\quad
\(
\inferrule{
  \const{C}( \vec t \,) \teq \const{D}( \vec u ) \in \Ec
  \betweenantes
  \const{C} \not= \const{D}
}{
  \bot
}
\)
\rn{Clash}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for bidirectional closure%.
}
\label{fig:cc-rules}
\end{figure}

\begin{figure}[t!]
\vspace*{+6pt} %% TYPESETTING HACK
\small
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
  t : \delta \in \Data
  \betweenantes
  \Val \smash{\ec{t}} = \MU x.\; u
  \betweenantes
  x \in \FV( u )
}{
  \bot
}
\)
\rn{Acyclic}
\\[5\jot]
\(
\inferrule{
 t, u : \delta \in \Codata
 \betweenantes
 \Val \smash{\ec{t}} \aequiv \Val \smash{\ec{u}}
}{
 \Ec := \Ec,\: t \teq u
}
\)
\rn{Unique}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for acyclicity and uniqueness%.
}
\label{fig:ab-rules}
\end{figure}

\begin{figure}[t!]
\vspace*{+6pt} %% TYPESETTING HACK
\small
\centering
\begin{tabular}{c}
\(
\inferrule{
  t : \delta
  \betweenantes
  t \in \tEc
  \betweenantes
  \Ctr^\delta = \{ \const{C}_1, \ldots, \const{C}_m \}
\\
  \bigl( \const s( t ) \in \tEc \text{ and } \const s \in \Sel^\delta \bigr)
  \text{ or }
  \bigl( \delta \in \Data \text{ and } %\left|
  \delta
  %\right|
  \text{ is finite} \bigr)
}{
  {\mathop{\big|\kern-.25ex\big|}\vthinspace}_{i=1}^n\, \Ec := \Ec,\: \DISC{i} %\ROR
 % \ROR \cdots \ROR \Ec := \Ec,\: \DISC{m}
}
\)
\rn{Split}
\\[5\jot]
\(
\inferrule{
  t, u : \delta \in \Codata
  \betweenantes
  t, u \in \tEc
  \betweenantes
  \delta \text{ is a singleton}
}{
  \Ec := \Ec,\: t \teq u
}
\)
\rn{Single}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for branching%.
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

\begin{examplex}
Suppose that $\Ec$ contains four distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{y}$, and~$\ec{z}$
such that $\const{C}(w,y) \in \ec{x}$ and $\const{C}(z,x) \in \ec{y}$ for some $\const{C} \in \Ctr$.
A possible sequence of unfolding steps is given below, omitting
trivial entries such as $\ec{w} \mapsto {\Varec{w}}$.
%
\begin{enumerate}
\item Unfold $\Varec{x}$:\enskip $\Val = \{\vthinspace  \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) \vthinspace\}$
\item Unfold $\Varec{y}$:\enskip $\Val = \{\vthinspace  \!\begin{aligned}[t]
  & \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \MU {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\; \\[-.5\jot]
  & \ec{y} \mapsto \MU {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) \vthinspace\}\end{aligned}$
\item Unfold $\Varec{x}$:\enskip $\Val = \{\vthinspace \!\begin{aligned}[t]
  & \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \MU {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\; \\[-.5\jot]
  & \ec{y} \mapsto \MU {\Varec{y}}.\; \const{C}( {\Varec{z}},\; \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) ) \vthinspace\}\end{aligned}$
\end{enumerate}
%
The resulting $\Val$ indicates that the values for $x$ and $y$ in models of $\Ec$
must be of the forms $\const{C}( {\Varec{w}}, \const{C}( {\Varec{z}},\allowbreak \const{C}( {\Varec{w}},\allowbreak \const{C}( {\Varec{z}},\ldots ))))$
and $\const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \ldots ))))$,
respectively. %, for some values of ${\Varec{w}}$ and ${\Varec{z}}$.
\xend
\end{examplex}

@@

\begin{examplex}
Let $\Ec = \{ x \teq \const{S}(x),\; y \teq \const{S}(\const{S}(y)) \}$.
After phase~1, the equivalence classes %of $\Ec$
are
$\{ x,\, \const{S}(x) \}$,
$\{ y,\, \const{S}(\const{S}(y)) \}$,
and $\{ \const{S}(y) \}$.
Constructing $\Val$ yields
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x}) 
\kern2.0em %% TYPESETTING
\Val \ec{y}  \,=\,
\MU {\Varec{y}}.\; \const{S}(\smash{\MU {\VAREC{\const{S}(y)}}}.\; \const{S}( {\Varec{y}} ) )
%\kern0.25em %% TYPESETTING
% &
\\
\Val \ec{\const{S}(y)} & \,=\, 
\MU {\VAREC{\const{S}(y)}}.\; \const{S}(\MU {\Varec{y}}.\; \const{S}( {\VAREC{\const{S}(y)}} ) )
\end{align*}
Since $\Val \ec{y} \aequiv \Val \ec{\const{S}(y)}$,
%for the renaming
%$\{ {\Varec{y}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{y}} \}$,
the \rn{Unique} rule applies to derive $y \teq \const{S}(y)$.
%Intuitively, this equality holds, since the values of $y$ and $\const{S}(y)$
%are equivalent.
%
At this point, phase~1 is activated again, % to XXX the equivalence classes,
yielding the equivalence classes
$\{ x,\, \const{S}(x) \}$ and
$\{ y,\, \const{S}(y),\allowbreak\, \const{S}(\const{S}(y)) \}$.
The mapping $\Val$ is updated accordingly:
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x})
&
\Val \ec{y} & \,=\, \MU {\Varec{y}}.\; \const{S}({\Varec{y}})
\end{align*}
Since $\Val \ec{x} \aequiv \Val \ec{y}$,
%for the renaming
%$\{ {\Varec{y}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{y}} \}$,
\rn{Unique} can finally be applied to derive $x \teq y$.
\xend
\end{examplex}

@@

>>>>

\newcommand\yasx{\mskip.5mu y\mskip.5mu} %%% TYPESETTING: hack so that y is exactly as wide as x

@@

Given a fixed signature, %~$\Sig$,
the decision procedure for the universal theory of (co)data\-types %$\thD$
determines the
%$\thD$-
satisfiability of finite sets $\Ec$ of literals:\ equalities
and disequalities between first-order terms, %$\Sig$-terms,
whose variables are interpreted existentially.
The decision procedure is formulated as an %tableau-like
abstract calculus.
%Following the tradition of refutational theorem proving,
Proving a universal quantifier-free conjecture is reduced to showing that
its negation is unsatisfiable. 
%The presentation is
%inspired by Barrett et al.\ \shortcite{barrett-et-al-2007} but at
%a higher level, using unoriented equations instead of oriented ones.

@@

\begin{figure}[b!]
%\small
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
  t \in \tEc
}{
  \Ec := \Ec,\: t \teq t
}
\)
\rn{Refl}
\quad
\(
\inferrule{
 t \teq u \in \Ec
}{
 \Ec := \Ec,\: u \teq t
}
\)
\rn{Sym}
\\[5\jot]
\(
\inferrule{
  s \teq t,\; t \teq u \in \Ec
}{
  \Ec := \Ec,\: s \teq u
}
\)
\rn{Trans}
\quad
\(
\inferrule{
  \vec t \teq \vec u \in \Ec
  \betweenantes
  \const f( \vec t \,),\, \const f( \vec u ) \in \tEc
}{
  \Ec := \Ec,\: \const f( \vec t \,) \teq \const f( \vec u )
}
\)
\rn{Cong}\\[5\jot]
\(
\inferrule{
  t \teq u,\; t \tneq u \in \Ec
}{
  \bot
}
\)
\rn{Conflict}
\quad
\(
\inferrule{
  \const{C}( \vec t \,) \teq \const{C}( \vec u ) \in \Ec
}{
  \Ec := \Ec,\: \vec t \teq \vec u
}
\)
\rn{Inject}
\\[5\jot]
\(
\inferrule{
  \const{C}( \vec t \,) \teq \const{D}( \vec u ) \in \Ec
  \betweenantes
  \const{C} \not= \const{D}
}{
  \bot
}
\)
\rn{Clash}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for bidirectional closure%.
}
\label{fig:cc-rules}
\end{figure}

@@

\begin{figure}[b!]
%\vspace*{+6pt} %% TYPESETTING HACK
%\small
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
  t : \delta \in \Data
  \betweenantes
  \Val \smash{\ec{t}} = \MU x.\; u
  \betweenantes
  x \in \FV( u )
}{
  \bot
}
\)
\rn{Acyclic}
\\[5\jot]
\(
\inferrule{
 t, u : \delta \in \Codata
 \betweenantes
 \Val \smash{\ec{t}} \aequiv \Val \smash{\ec{u}}
}{
 \Ec := \Ec,\: t \teq u
}
\)
\rn{Unique}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for acyclicity and uniqueness%.
}
\label{fig:ab-rules}
\end{figure}

@@

\begin{examplex}
Suppose that $\Ec$ contains four distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{\yasx}$, and~$\ec{z}$
such that $\const{C}(w,y) \in \ec{x}$ and $\const{C}(z,x) \in \ec{\yasx}$ for some $\const{C} \in \Ctr$.
A possible sequence of unfolding steps is given below, omitting
trivial entries such as $\ec{w} \mapsto {\Varec{w}}$.
%
\begin{enumerate}
\item Unfold $\Varec{x}$:\enskip $\Val = \{\vthinspace  \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{\yasx}} ) \vthinspace\}$
\item Unfold $\Varec{\yasx}$:\enskip $\Val = \{\vthinspace  \!\begin{aligned}[t]
  & \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \MU {\Varec{\yasx}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\; \\[-.5\jot]
  & \ec{\yasx} \mapsto \MU {\Varec{\yasx}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) \vthinspace\}\end{aligned}$
\item Unfold $\Varec{x}$:\enskip $\Val = \{\vthinspace \!\begin{aligned}[t]
  & \ec{x} \mapsto \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \MU {\Varec{\yasx}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\; \\[-.5\jot]
  & \ec{\yasx} \mapsto \MU {\Varec{\yasx}}.\; \const{C}( {\Varec{z}},\; \MU {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{\yasx}} ) ) \vthinspace\}\end{aligned}$
\end{enumerate}
%
The resulting $\Val$ indicates that the values for $x$ and $y$ in models of $\Ec$
must be of the forms $\const{C}( {\Varec{w}}, \const{C}( {\Varec{z}},\allowbreak \const{C}( {\Varec{w}},\allowbreak \const{C}( {\Varec{z}},\ldots ))))$
and $\const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \ldots ))))$,
respectively. %, for some values of ${\Varec{w}}$ and ${\Varec{z}}$.
\xend
\end{examplex}

@@

\begin{examplex}
Let $\Ec = \{ x \teq \const{S}(x),\; y \teq \const{S}(\const{S}(y)) \}$.
After phase~1, the equivalence classes %of $\Ec$
are
$\{ x,\, \const{S}(x) \}$,
$\{ y,\, \const{S}(\const{S}(y)) \}$,
and $\{ \const{S}(y) \}$.
Constructing $\Val$ yields
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x}) 
\kern2.0em %% TYPESETTING
\Val \ec{\yasx}  \,=\,
\MU {\Varec{\yasx}}.\; \const{S}(\smash{\MU {\VAREC{\const{S}(y)}}}.\; \const{S}( {\Varec{\yasx}} ) )
%\kern0.25em %% TYPESETTING
% &
\\
\Val \ec{\const{S}(y)} & \,=\, 
\MU {\VAREC{\const{S}(y)}}.\; \const{S}(\MU {\Varec{\yasx}}.\; \const{S}( {\VAREC{\const{S}(y)}} ) )
\end{align*}
Since $\Val \ec{\yasx} \aequiv \Val \ec{\const{S}(y)}$,
%for the renaming
%$\{ {\Varec{\yasx}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{\yasx}} \}$,
the \rn{Unique} rule applies to derive $y \teq \const{S}(y)$.
%Intuitively, this equality holds, since the values of $y$ and $\const{S}(y)$
%are equivalent.
%
At this point, phase~1 is activated again, % to XXX the equivalence classes,
yielding the equivalence classes
$\{ x,\, \const{S}(x) \}$ and
$\{ y,\, \const{S}(y),\allowbreak\, \const{S}(\const{S}(y)) \}$.
The mapping $\Val$ is updated accordingly:
\begin{align*}
\Val \ec{x} & \,=\, \MU {\Varec{x}}.\; \const{S}(\Varec{x})
&
\Val \ec{\yasx} & \,=\, \MU {\Varec{\yasx}}.\; \const{S}({\Varec{\yasx}})
\end{align*}
Since $\Val \ec{x} \aequiv \Val \ec{\yasx}$,
%for the renaming
%$\{ {\Varec{\yasx}} \mapsto {\VAREC{\const{S}(y)}}, {\VAREC{\const{S}(y)}} \mapsto {\Varec{\yasx}} \}$,
\rn{Unique} can finally be applied to derive $x \teq y$.
\xend
\end{examplex}

@@

\begin{figure}[t!]
%\vspace*{+6pt} %% TYPESETTING HACK
%\small
\centering
\begin{tabular}{@{}c@{}}
\(
\inferrule{
  t : \delta
  \betweenantes
  t \in \tEc
  \betweenantes
  \Ctr^\delta = \{ \const{C}_1, \ldots, \const{C}_m \}
\\
  \bigl( \const s( t ) \in \tEc \text{ and } \const s \in \Sel^\delta \bigr)
  \text{ or }
  \bigl( \delta \in \Data \text{ and } %\left|
  \delta
  %\right|
  \text{ is finite} \bigr)
}{
  {\mathop{\big|\kern-.25ex\big|}\vthinspace}_{i=1}^n\, \Ec := \Ec,\: \DISC{i} %\ROR
 % \ROR \cdots \ROR \Ec := \Ec,\: \DISC{m}
}
\)
\rn{Split}
\\[5\jot]
\(
\inferrule{
  t, u : \delta \in \Codata
  \betweenantes
  t, u \in \tEc
  \betweenantes
  \delta \text{ is a singleton}
}{
  \Ec := \Ec,\: t \teq u
}
\)
\rn{Single}
\end{tabular}
\vspace*{-2pt} %% TYPESETTING HACK
\caption{\,Derivation rules for branching%.
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}
